Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-03-07 docs/papers/roofline_model/gocean_roofline.tex
roofline: add more discussion of Continuity SIMD results.

In contrast to the results obtained with the Shallow kernel, the
performance of the NEMOLite2D u-Momentum kernel shows little
sensitivity to working-set size -- the $128^2$, $256^2$ and $512^2$
performance points in Figure~\ref{FIG_roofline_archer} are very close
together.

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1.1~MB and 4.7~MB, respectively and
therefore both of these should fit comfortably within L3
cache. Despite this, the performance achieved with these domain sizes
falls short of the L3 bandwidth limit.  However, the Continuity kernel is just
one of several within the time-stepping loop of NEMOLite2D; the loop
as a whole has a working-set size of $180\times n^{2}$ bytes. For the
$32^2$ domain this is 180~KB, for $52^2$ it is 475~KB and for the
$128^2$ domain it is 2.9~MB which is well within the capacity of L3.

@@

The continuity kernel includes array accesses of the form {\tt
  a(i,j-1)}. If array {\tt a} is allocated with extent {\tt (Nx, Ny)}
then this specifies a location in memory {\tt Nx} words before that
occupied by {\tt a(i,j)}. In order for such an access to be aligned,
{\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
AVX) and this must be specified to the compiler. We therefore padded Nx
as required and added a directive within the wrapper routine:
\begin{verbatim}
!DIR$ ASSUME (MOD(Nx,ALIGNMENT) .EQ. 0)
\end{verbatim}

Once all of these steps were complete, the only un-aligned array
accesses reported by the compiler were those of the form {\tt
  a(i-1,j)}. Without taking a physical copy of the array {\tt a}, such
unaligned accesses are unavoidable in a stencil-based
finite-difference code.

>>>>

In contrast to the results obtained with the Shallow kernel, the
performance of the NEMOLite2D u-Momentum kernel shows little
sensitivity to working-set size -- the $128^2$, $256^2$ and $512^2$
performance points in Figure~\ref{FIG_roofline_archer} are very close
together and well below any of the roofline limits.

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1.1~MB and 4.7~MB, respectively and
therefore both of these should fit comfortably within L3
cache. Despite this, the performance achieved with these domain sizes
falls short of the L3 bandwidth limit.  Even when we consider that the
Continuity kernel is just one of several within the time-stepping loop
of NEMOLite2D we see that the L3 bandwidth limit is still applicable;
the loop as a whole has a working-set size of $180\times n^{2}$
bytes. For the $32^2$ domain this is 180~KB ({\it i.e.}\ within L2),
while the $52^2$ (475~KB) through to the $128^2$ domain (2.9~MB) fall
well within the capacity of L3.

@@

The continuity kernel includes array accesses of the form {\tt
  a(i,j-1)}. If array {\tt a} is allocated with extent {\tt (Nx, Ny)}
then this specifies a location in memory {\tt Nx} words before that
occupied by {\tt a(i,j)}. In order for such an access to be aligned,
{\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
AVX) and this must be specified to the compiler. We therefore padded Nx
as required and added a directive within the wrapper routine:
\begin{verbatim}
!DIR$ ASSUME (MOD(Nx,ALIGNMENT) .EQ. 0)
\end{verbatim}
where {\tt ALIGNMENT} is an integer parameter declared to have the value two
when compiling with SSE and four with AVX.

We can see in Figure~\ref{FIG_cont_vec} that these changes (``SSE
align'') significantly improve the performance of the smaller
domains. Again, as the domain size increases this improvement reduces
until, for $512^2$ and $1024^2$ (where we are limited by the bandwidth
to main memory), this optimisation has no effect.

Since the Continuity kernel is one of several in the time-stepping
loop, we experimented with pre-loading the fields into cache simply by
executing the Continuity kernel immediately prior to executing it for
a second time with timing. Figure~\ref{FIG_cont_vec} shows that this
``pre-loading'' slightly improved the kernel performance for those
problem sizes that fit within cache. {\bf Discuss this}

Once all of the above steps were complete, the only un-aligned array
accesses reported by the compiler were those of the form {\tt
  a(i-1,j)}. Without taking a physical copy of the array {\tt a}, such
unaligned accesses are unavoidable in a stencil-based
finite-difference code. However, in order to quantify how much these
accesses limit the performance of the kernel we experimented with
replacing e.g. {\tt a(i-1,j)} with {\tt a(i-ALIGNMENT,j)}. The
resulting kernel had no un-aligned accesses and performed
significantly better; actually reaching or exceeding the L3-bandwidth
roofline for domains $64^2$ -- $256^2$.
